Number of COVID-19 Cases in Canada By Region, Age, and Sex

### 2024-11-26 paper/paper.qmd
Updated prose of paper

The COVID-19 pandemic is something that is difficult to forget. Although it is now in the past, the effects that it had on society and how we function as individuals is still profound. At the same time, the statistical information that sprouted from the pandemic has a high value for analysis. This is not only to better understand the pandemic and how it affected populations but also to form a strong idea of its patterns to suggest tactics for dealing with pandemics shall something similar happen in the future.

Studies have been done in recent years in attempts to utilize the data that was gathered during COVID times to better advise future preventions and treatment for pandemic situations. Many studies look into the relationship between demographic information of populations and their likelihood of getting diagnosed or fatality. One study found that communities on hillsides and other locations with enhanced ventilation and oxygen had reduced numbers of COVID cases [@cite-covid_demographics]. Other studies have found relationships between exercise and cardiovascular function and the likelihood of getting COVID [@cite-cardio_and_covid]. Although many relationships have been examined, there has been a lack of studies focusing on a broad set of demographic traits such as age and sex. There has also not been cross comparisons of region and demographic aspects. Another gap is the lack of country-specific analysis, particularly Canada.

@@

The analysis revealed that there are disparities in the risk of severe COVID-19 outcomes across demographic and regional categories. Logistic regression models demonstrated that individuals in rural regions consistently faced higher probabilities of being classified as high risk, with males and specific age intervals further amplifying this vulnerability. These findings were supported by consistent trends in summary statistics and model visualizations, which revealed the interplay between predictors such as age, region, and sex in influencing COVID-19 outcomes.

Understanding the drivers of COVID-19 severity can craft equitable and effective public health responses. By identifying which demographic and regional groups face the greatest risks, the results of the analysis can guide resource distribution, including vaccination programs and medical support, to areas where they are most needed. These are applicable not only to COVID-19 but also to future public health challenges.

@@

The dataset reflects a transformation of real-world phenomena—COVID-19 cases, deaths, and testing rates—into structured entries that can be analyzed. Each entry in the dataset represents aggregated information collected from health departments, governmental organizations, and research institutions worldwide. These organizations report COVID-19 statistics at varying levels of granularity, such as daily case counts or weekly summaries, which are then standardized and compiled into the dataset. This collective and comprehensive effort of data collection allows the real-world challenges of facing COVID-19 to be transformed into quantitative measures that can be analyzed and modeled.

@@

Another key advantage of the logistic regression model in this context is that it provides interpretable coefficients, which allow for an understanding of how specific predictors influence the odds of being high risk. This interpretability is crucial for public health applications, where stakeholders need clear and actionable insights. Furthermore, logistic regression is computationally efficient, making it ideal for datasets with relatively straightforward structures like this one. Thus, the logistic regression model strikes an appropriate balance between accuracy, interpretability, and simplicity, making it the most suitable choice for this analysis.

@@

One key limitation of logistic regression is its inability to account for nuanced interactions between predictors without explicitly adding interaction terms. For instance, the combined effects of age and sex on risk may not be fully captured in this model. Additionally, the model assumes a predefined functional form, which limits its flexibility in identifying nonlinear relationships between predictors and outcomes. As a result, more complex patterns in the data may remain undetected, suggesting the need for alternative modeling approaches such as random forests or decision trees in future studies.

@@

A confusion matrix was also constructed to assess classification accuracy, precision, and recall by comparing the predicted classes to the actual outcomes. Values of above 0.75 were yielded for the accuracy rating. Lastly, the odds ratios with confidence intervals provided insights into the strngth and direction of the predictors' effects.

@@

The results of this analysis reveal critical insights into how demographic and regional factors influence the likelihood of being classified as high risk for COVID-19. The logistic regression model, designed to predict the probability of high risk, effectively highlighted the relative importance of age, sex, and region in shaping these probabilities. Notably, rural regions consistently showed higher predicted probabilities of high risk compared to urban regions. This disparity underscores the challenges faced by rural areas, such as limited access to healthcare infrastructure, which may contribute to higher case fatality rates (CFR).

Similarly, the analysis revealed that males are more likely to be classified as high risk than females, aligning with previous research suggesting that biological differences, comorbidities, and health-seeking behaviors could account for this discrepancy. Age also played a pivotal role, with certain age intervals, such as 10–20 and 30–40, showing significantly higher probabilities of high risk. This pattern may reflect complex interactions between behavioral, immunological, and societal factors influencing COVID-19 outcomes. Understanding these relationships is crucial for tailoring public health interventions and resource allocation strategies.

@@

Building on the findings of this study, future research should explore several directions. First, expanding the analysis to include additional countries or regions would provide a more comprehensive understanding of the factors influencing COVID-19 outcomes. Second, incorporating additional variables, such as socioeconomic status, vaccination rates, and comorbidities, could enhance the predictive power of the models.

Another next step that could be taken is to use the analysis and modeling on a broader series of pandemics and diseases such as seasonal flues, SARS outbreak in the 2000s, and others that may show similar trends. This will allow a broader generalizability of the demographic patterns and allow the model to better predict probability of high risk based on demographic data. Finally, public health practitioners and researchers should collaborate to ensure that findings are translated into actionable policies and interventions, bridging the gap between data analysis and real-world impact.

@@

The dataset used in this analysis is sourced from the COVerAGE-DB, a comprehensive repository of COVID-19 demographic data. The database compiles data from a diverse set of sources, including government health agencies, international organizations, and other reputable institutions, ensuring a broad representation of regions and demographics. Each entry in the dataset represents an aggregation of cases, deaths, and demographic information categorized by age, sex, and region.

>>>>

The COVID-19 pandemic is something that is difficult to forget. Although it is now in the past, the effects that it had on society and how we function as individuals is still noticeable. At the same time, the statistical information that sprouted from the pandemic has a high value for analysis. This is not only to better understand the pandemic and how it affected populations but also to form a strong idea of its patterns to suggest tactics for dealing with pandemics shall something similar happen in the future.

Studies have been done in recent years in attempts to utilize the data that was gathered during COVID times to better advise future preventions and treatment for pandemic situations. Many studies look into the relationship between demographic information of populations and their likelihood of getting diagnosed or fatality. One study found that communities on hillsides and other locations with enhanced ventilation and oxygen had reduced numbers of COVID cases [@cite-covid_demographics]. Other studies have found relationships between exercise and cardiovascular function and the likelihood of getting COVID [@cite-cardio_and_covid]. Although many relationships have been examined, there has been a lack of studies focusing on a broad set of demographic traits such as age and sex. There has also not been cross comparisons of region and demographic aspects. Another area where research is needed is for country-specific analysis, particularly Canada.

@@

The analysis showed that there are disparities in the risk of severe COVID-19 outcomes across demographic and regional categories. Logistic regression models demonstrated that individuals in rural regions consistently faced higher probabilities of being classified as high risk, with males and specific age intervals further amplifying this vulnerability. These findings were supported by consistent trends in summary statistics and model visualizations, which showed the interplay between predictors such as age, region, and sex in influencing COVID-19 outcomes.

Understanding the reasons behind COVID-19 severity can craft equitable and effective public health responses. By identifying which demographic and regional groups face the greatest risks, the results of the analysis can guide resource distribution, including vaccination programs and medical support, to areas where they are most needed. These are applicable not only to COVID-19 but also to future public health challenges.

@@

The dataset reflects a transformation of real-world phenomena—COVID-19 cases, deaths, and testing rates—into structured entries that can be analyzed. Each entry in the dataset represents aggregated information collected from health departments, governmental organizations, and research institutions worldwide. These organizations report COVID-19 statistics at varying levels of granularity, such as daily case counts or weekly summaries, which are then standardized and compiled into the dataset. This effort of data collection allows the real-world challenges of facing COVID-19 to be transformed into quantitative measures that can be analyzed and modeled.

@@

Another key advantage of the logistic regression model in this context is that it provides interpretable coefficients, which allow for an understanding of how specific predictors influence the odds of being high risk. This interpretability is crucial for public health applications. Furthermore, logistic regression is computationally efficient, making it ideal for datasets with relatively straightforward structures like this one. Thus, the logistic regression model strikes an appropriate balance between accuracy, interpretability, and simplicity, making it the most suitable choice for this analysis.

@@

One key limitation of logistic regression is its inability to account for interactions between predictors without explicitly adding interaction terms. For instance, the combined effects of age and sex on risk may not be fully captured in this model. Additionally, the model assumes a predefined functional form, which limits its flexibility in identifying nonlinear relationships between predictors and outcomes. As a result, more complex patterns in the data may remain undetected, suggesting the need for alternative modeling approaches such as random forests or decision trees in future studies.

@@

A confusion matrix was also constructed to assess classification accuracy, precision, and recall by comparing the predicted classes to the actual outcomes. Values of above 0.75 were yielded for the accuracy rating. Lastly, the odds ratios with confidence intervals provided information on the strength and direction of the predictors' effects.

@@

The results of this analysis displays information into how demographic and regional factors influence the likelihood of being classified as high risk for COVID-19. The logistic regression model, designed to predict the probability of high risk, effectively highlighted the relative importance of age, sex, and region in shaping these probabilities. Notably, rural regions consistently showed higher predicted probabilities of high risk compared to urban regions. This disparity underscores the challenges faced by rural areas, such as limited access to healthcare infrastructure, which may contribute to higher case fatality rates (CFR).

Similarly, the analysis revealed that males are more likely to be classified as high risk than females, aligning with previous research suggesting that biological differences, comorbidities, and health-seeking behaviors could account for this discrepancy. Age also played a pivotal role, with certain age intervals, such as 10–20 and 30–40, showing significantly higher probabilities of high risk. This pattern may reflect complex interactions between behavioral, immunological, and societal factors influencing COVID-19 outcomes. Understanding these relationships is important for tailoring public health interventions and resource allocation strategies.

@@

Building on the findings of this study, future research should explore several directions. First, expanding the analysis to include additional countries or regions would provide a better understanding of the factors influencing COVID-19 outcomes. Second, incorporating additional variables, such as socioeconomic status, vaccination rates, and comorbidities, could enhance the predictive power of the models.

Another next step that could be taken is to use the analysis and modeling on a broader series of pandemics and diseases such as seasonal flues, SARS outbreak in the 2000s, and others that may show similar trends. This will allow a broader generalizability of the demographic patterns and allow the model to better predict probability of high risk based on demographic data. Finally, public health practitioners and researchers should collaborate to ensure that findings are translated into actionable policies and interventions.

@@

The dataset used in this analysis is sourced from the COVerAGE-DB, a repository of COVID-19 demographic data. The database compiles data from a diverse set of sources, including government health agencies, international organizations, and other reputable institutions, ensuring a broad representation of regions and demographics. Each entry in the dataset represents an aggregation of cases, deaths, and demographic information categorized by age, sex, and region.
